---
title: "Module 2: Post-Quiz Code Document"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Lesson 1

In case you had difficulty with coming up with the code for the practice quiz,
the correct code is provided here. This code will be provided to you after each
practice quiz in the module and will serve as a resource for when you’re stuck.
It will remain locked until you first attempt the practice quiz, and the code
will be updated each time to reflect the progress made on the Indicator Table up
to that point.

In preparation for reproducing the Indicator Table from the Global Findex
database report, the data first needs to be saved as an object in your global
environment.

```{r, message = FALSE}
# read in data
micro_world <- read_csv("micro_world.csv")
```

# Lesson 2

*Previous Progress*

```{r, message = FALSE}
# read in data
micro_world <- read_csv("micro_world.csv")
```

By this point of Week 2, you should now have created a data frame of the overall
percentage of adults with an account by country, as well as a data frame of the
percentage of adults with an account by both country and gender.

```{r, message = FALSE}
# overall account ownership
overall_account <- 
  micro_world %>%
  group_by(economy) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt))

# account ownership by gender
account_gap_female <- 
  micro_world %>%
  group_by(economy, female) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt))
```

# Lesson 3

*Previous Progress*

```{r, message = FALSE}
# read in data
micro_world <- read_csv("micro_world.csv")

# overall account ownership
overall_account <- 
  micro_world %>%
  group_by(economy) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt))

# account ownership by gender
account_gap_female <- 
  micro_world %>%
  group_by(economy, female) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt))
```

Building on the previous assessment, you should now have the code for obtaining
both the gender gap and income gap in account ownership.

```{r, message = FALSE}
# gender gap in account ownership
account_gap_female <- 
  micro_world %>%
  group_by(economy, female) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt)) %>%
  pivot_wider(id_cols = economy, 
              names_from = female, 
              names_prefix = "female",
              values_from = account) %>%
  mutate(gap_female = female1 - female2) %>%
  select(economy, gap_female)

# income gap in account ownership
account_gap_inc <- 
  micro_world %>%
  mutate(lower_inc = inc_q <= 2) %>%
  group_by(economy, lower_inc) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt)) %>%
  pivot_wider(id_cols = economy, 
              names_from = lower_inc, 
              names_prefix = "lower_inc",
              values_from = account) %>%
  mutate(gap_inc = lower_incFALSE - lower_incTRUE) %>%
  select(economy, gap_inc)
```

# Lesson 4

*Previous Progress*

```{r, message = FALSE}
# read in data
micro_world <- read_csv("micro_world.csv")

# overall account ownership
overall_account <- 
  micro_world %>%
  group_by(economy) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt))

# gender gap in account ownership
account_gap_female <- 
  micro_world %>%
  group_by(economy, female) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt)) %>%
  pivot_wider(id_cols = economy, 
              names_from = female, 
              names_prefix = "female",
              values_from = account) %>%
  mutate(gap_female = female1 - female2) %>%
  select(economy, gap_female)

# income gap in account ownership
account_gap_inc <- 
  micro_world %>%
  mutate(lower_inc = inc_q <= 2) %>%
  group_by(economy, lower_inc) %>% 
  summarize(account = 100 * weighted.mean(account, w = wgt)) %>%
  pivot_wider(id_cols = economy, 
              names_from = lower_inc, 
              names_prefix = "lower_inc",
              values_from = account) %>%
  mutate(gap_inc = lower_incFALSE - lower_incTRUE) %>%
  select(economy, gap_inc)
```

Finally, you should now have the code necessary to reproduce the Indicator Table
by joining the data frames that you've created in the two previous assessments.

```{r, message = FALSE}
# indicator table
indicator_table <- 
  full_join(overall_account, 
            account_gap_female, 
            by = "economy") %>%
  full_join(account_gap_inc, 
            by = "economy") %>%
  mutate(account = round(account),
         gap_female = round(gap_female),
         gap_inc = round(gap_inc))
```

If you are interested in creating an auxliary table that shows only the
statistically significant gaps, you will first need to calculate the margin of
error for each country as defined in the footnote of Table A.1 from the Global
Findex database report. The gaps that are not statistically significant (i.e.,
gaps falling within the margin of error) are replaced with a dash (--), and the
`case_when` function is used to accomplish this task. Consider exploring the
help page of the `case_when` function to familiarize yourself with its usage.  

```{r, message = FALSE}
# margin of error
moe <- 
  micro_world %>% 
  group_by(economy, pop_adult) %>% 
  summarize(design_effect = n() * sum(wgt^2) / (sum(wgt)^2),
            moe = 100 * sqrt(0.25 / n()) * 1.96 * sqrt(design_effect))

# auxiliary table
aux_indicator_table <- 
  full_join(overall_account, 
            account_gap_female, 
            by = "economy") %>%
  full_join(account_gap_inc, 
            by = "economy")  %>%
  full_join(moe, 
            by = "economy") %>%
  mutate(account = round(account), 
         gap_female = 
           case_when(
             abs(gap_female) > moe ~ as.character(round(gap_female)), 
             TRUE ~ "-"
           ),
         gap_inc = 
           case_when(
             abs(gap_inc) > moe ~ as.character(round(gap_inc)), 
             TRUE ~ "-"
           )) %>% 
  select(economy, account, gap_female, gap_inc)
```
