---
title: "Module 3, Lesson 2: Collective geoms"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the second lesson of Module 3, you learned about histograms, barplots, and
boxplots -- all of which are "collective geoms". Remember that collective geoms are 
so-called because rather than plotting individual data they plot summaries of the
data. A collective geom is intended to show some overall feature of the data
instead of specific data points. 

Open the R project called `Lessons.Rproj` and a blank R script. Then, 
load the necessary packages into the current R session:

```{r}
library(healthds)
library(tidyverse)
```


## `geom_histogram`

As you saw in the video, a histogram is created by first binning or grouping
together observations having numeric values that are considered near to one another
in some sense (more on this in a moment) and then plotting these groups as rectangles 
with height equal to the number of observations in a bin.

To create a histogram of the infant weights, we use the following code, mapping `weight_kg` to the `x` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 1)
```

From the plotted result, we see that two measured weights were between 
2.5 and 3.5, because the height of the bar is 2. Similarly, two were between 
3.5 and 4.5,3 were between 4.5 and 5.5, and one was between 6.5 and 7.5. By
constructing a histogram, we exchange precision of individual measurements (e.g. 
we can't say anything more than that a measurement feel *somewhere* between 
6.5kg and 7.5kg) for a better characterization of the whole distribution of 
weight measurements.

This geom has an extra argument that we've not seen before, `binwidth`, 
which we set equal to 1 in the call above. Note that it falls *outside* of
the aesthetic mappings in the `aes` function. This choice means that the
size of the bins is equal to 1. For example, the first bin starts at 2.5
and ends at 3.5, hence the width is 3.5 - 2.5 or 1. And the same goes for
the rest of the bins: 4.5 - 3.5, 5.5 - 4.5, etc. Different choices of `binwidth` 
may give dramatically different results, particularly for small datasets like 
this one. Try out different values of `binwidth` to see for yourself. 

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 0.25)
```

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 2.0)
```

You might be thinking that only choosing `binwidth` is not sufficient to determine
how the histogram should appear. For example, if `binwidth` is 1, you could have
a bin of (2.5, 3.5) or (2.6, 3.6) or  (2.7, 3.7) or... How does `ggplot` 
choose (2.5, 3.5)? In fact, if you only specify `binwidth`, by default `ggplot` will 
anchor all bins based upon the (potentially theoretical) bin centered at 0, 
with boundaries equal to half of `binwidth` in either direction (I say this is 
"potentially theoretical" because your numbers may be much much bigger than 0). 
So this anchoring bin in our case is (-0.5, 0.5). Once you know this bin, all 
of the other bins become automatically determined, i.e. (0.5, 1.5), (1.5, 2.5), etc.

As you might be figuring out by know, you can override this default behavior
by specifying `center`, as in:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 1, 
                 center = 1.5)
```

Centering the first bin at 1.5, the bins become (2, 3), (3, 4), etc.
Alternatively, instead of specifying `center`, you can specify `boundary`,
which gives the edge of a bin:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 1, 
                 boundary = 1)
```

Setting `center = 0.5` gives the same result as `boundary = 1` (when `binwidth=1`). 

There are opportunities to have your histogram tell a different story through 
your choice of `binwidth` and `center` or `boundary`. For example, consider what
happens when we move the value of `center` from 1.5 above to 1.4 below:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg), 
                 binwidth = 1, 
                 center = 1.4)
```

A small change in one argument results in a much different-looking histogram.
Related to this, if you fail to choose `binwidth`, `ggplot` will make some default decisions, and these won't necessarily be good decisions:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(x = weight_kg))
```

Having failed to set `binwidth`, `ggplot` informs us of this fact and suggests to choose our own value. 

Your responsibility as a data scientist is to choose values that most
faithfully represent your data, and this can be challenging sometimes!

Lastly, note that you can create a horizontal histogram by mapping your 
variable to the `y` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_histogram(mapping = aes(y = weight_kg), 
                 binwidth = 1)
```

Now, the x-axis is used to represent the counts. 

Save this script as `03_02collectivegeoms.R`.

## `geom_bar`

A barplot, created using `geom_bar`, is similar to a histogram, the main difference 
being that it is intended to be used for *characters* rather than *doubles* (the
Independent Guide for Module 3, Lesson 1 also talks about *characters* versus *doubles*).
A simple barplot counting the number of observations per infant id is created by:

```{r}
ggplot(data = growth_data) + 
  geom_bar(mapping = aes(x = id))
```

Barplots, like histograms, give counts. And from this barplot we see that
there are four observations per id. 

Often, one wants to incorporate more aesthetics into a barplot. To demonstrate, 
we will apply some of our `dplyr` knowledge to add a new variable to our 
`growth_data`. Specifically, we will create a variable that equals `TRUE` when 
a measured weight exceeds 5kg and `FALSE` otherwise. Different from the guided 
practice video, we will create this new variable and *assign* the resulting 
data frame to a new dataset called `growth_data_expanded` (we used this same
variable name in Module 2, Lesson 3; although the names are the same, they 
are unrelated):

```{r}
growth_data_expanded <- 
  growth_data %>%
  mutate(weight_gt_5kg = weight_kg > 5)
```

The expression `weight_gt_5kg = weight_kg > 5` may be confusing at first glance. 
Remember from Module 2, Lesson 2 that the `=` sign is serving as a declaration, 
and so we are telling R to create a new variable called `weight_gt_5kg` that
takes on value equal to the right-hand side of the equation sign. Focusing now 
on the right-hand side, `weight_kg > 5` is asking a question: "is `weight_kg` 
greater than 5?" And the answer to this question is the value that `weight_gt_5kg` 
takes on. This happens on a row-by-row basis, so the answer to the question may 
be `FALSE` ('no') for one row and `TRUE` ('yes') for another row. 

By assigning this expanded dataset to a new object called `growth_data_expanded`, we 
preserve this result for future analyses. 

Returning to barplots, let's create a barplot that maps this `weight_gt_5kg` 
variable to the `fill` aesthetic:

```{r}
ggplot(data = growth_data_expanded) + 
  geom_bar(mapping = aes(x = id, fill = weight_gt_5kg))
```

Notice how, in contrast to the guided practice, we aren't piping the data into
`ggplot` here, but we could:

```{r, eval = FALSE}
growth_data_expanded %>% 
  ggplot() + 
  geom_bar(mapping = aes(x = id, fill = weight_gt_5kg))
```

(remember that the the pipe operator sends the thing that is being piped as the 
first argument of the function that the pipe points to, which in the case of the
function `ggplot` is `data`)

We have the same barplot as before but now with each bar filled to represent the number of counts for which `weight_gt_5kg` is `TRUE` and `FALSE`. These colored bars are, by default, stacked on top of each other. We can use the `position` argument inside of `geom_bar` change this behavior. For example, to stack the bars next to each other, we do:

```{r}
ggplot(data = growth_data_expanded) + 
  geom_bar(mapping = aes(x = id, fill = weight_gt_5kg), 
           position = "dodge")
```

Each value if the `id` variable now gets *two* bars, one for `weight_gt_5kg` equal 
to `TRUE` and another for `weight_gt_5kg` equal to `FALSE`. We are abusing the
`x` aesthetic slightly here -- the fact that the reddish bar corresponding to
`FALSE` is on the left and greenish bar corresponding to `TRUE` is on the
right should not be interpreted to mean that they have different values of `id`. 
Dodging the bars this way allows for an easier direct comparison of the colored 
bars to one another, at the cost of making it slightly more difficult to compare 
the total counts combining all colors. 

We can also normalize the heights of the bars so that the filled bars represent 
proportions instead of counts:

```{r}
ggplot(data = growth_data_expanded) + 
  geom_bar(mapping = aes(x = id, fill = weight_gt_5kg), 
           position = "fill")
```

(unfortunately, there are two distinct uses of the word 'fill' here. It is an 
aesthetic that can be mapped inside of the `aes()` function and also a word, in quotes, 
that specifies a type of barplot.) Although setting `position="fill"` doesn't 
look different at first glance from the default stacked barplot, this is only the 
case because each value of `id` has equal numbers of observations. Upon closer 
inspection one notices that the y-axis ends at 1 instead of at 4 in the stacked barplot. 
All bars will always have height equal to 1 when using `position="fill"`. Because
of this, the default label for the y-axis ("count") is no longer accurate, and 
you can change this label using what you will learn in Module 3, Lesson 3.

One potential source of confusion is understanding the difference between the 
`color` and `fill` aesthetics. Generally speaking, if a geometric object has 
area to it (like a bar), the `fill` aesthetic controls the shading of the inside 
and the `color` aesthetic controls the shading of its perimeter. If you're not 
sure which is appropriate, try them both and see which you prefer:

```{r}
ggplot(data = growth_data_expanded) + 
  geom_bar(mapping = aes(x = id, color = weight_gt_5kg))
```

In this case, coloring just the perimeter of the bars makes it hard to visually 
distinguish between groups, and we'd probably prefer to use the `fill` aesthetic.

Save your script to preserve the additional code you've added to this script.

## `geom_boxplot`

Much of what we've done so far has been within the `tidyverse` family of packages. 
In this lesson we briefly leave the `tidyverse` to learn about some new base
R functions: `sort` and `quantile`. We also learn about the `pull` function (which
happens to be in the `tidyverse` although it's new to us). We introduce these
functions only to help understand how `geom_boxplot` works. 

`pull` is analogous to `select` (Module 2, Lesson 3) but with two key differences:
it can only include one variable (whereas `select` can include more than one), and
it gives back a vector of numbers rather than a dataset. The advantage to `pull`ing
a variable is that we can then use the `sort` and `quantile` functions, which 
both only work on a vector of numbers. Consider the following three lines of code:

```{r}
growth_data %>% pull(weight_kg)
growth_data %>% pull(weight_kg) %>% sort()
growth_data %>% pull(weight_kg) %>% quantile()
```
The first pulls the `weight_kg` variable. From it, you can see the resulting vector. 
It is still comprised of the measured infant weights, but we have lost all context, 
e.g. we don't know which id or month each measured weight corresponds to.
The second line pulls this variable and then sorts it in increasing order. This should
should remind you of the `arrange` function (but remember that `arrange` is intended
to be used in the context of a dataset, not a vector). The third line pulls this variable
and then calculates the minimum, three quartiles, and maximum of this vector. 

The first, second, and third quartiles of a set of numbers are those for which 
25%, 50%, and 75%, respectively, of the numbers in the set fall below. These quartiles
form the *box* of a boxplot:

```{r}
ggplot(data = growth_data) + 
    geom_boxplot(mapping = aes(x = weight_kg))
```

The thick middle line is the second quartile (also called the median), and the
left and right edges of the box are the first and third quartiles. From the
boxplot, we know that 50% of the measured infant weights fall inside the box. The
lines on each end of the boxplot are sometimes called 'whiskers'. As discussed
in the videos, the whiskers encompass all data points within 1.5 times the 
interquartile range (the difference between the third and first quartiles) of the
box's edge. 

**You don't need to use the `quantile` function to create a boxplot.** All calculations
are done automatically simply by calling `geom_boxplot`. 

As with histograms and barplots, you can rotate a boxplot by mapping `weight_kg`
to the `y` aesthetic:

```{r}
ggplot(data = growth_data) + 
    geom_boxplot(mapping = aes(y = weight_kg))
```

Finally, you can create separate boxplots by mapping a categorical variable, such
as `id`, to another aesthetic, such as `y`:

```{r}
ggplot(data = growth_data) + 
    geom_boxplot(mapping = aes(x = weight_kg, y = id))
```

Save your script once again before closing.