---
title: "Introduction to blood pressure data"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In this course, we will analyze synthetic data that mimics -- but is distinct
from -- actual data reported in a study by @bhadoria.jfcm. We created these
data rather than using real data for simplicity and ease of logistics.

@bhadoria.jfcm invited 939 adults age 20 and older across 48 villages and 15
urban wards from Jabalpur district in central India to respond to a survey about
hypertension and cardiovascular risk factors. Nine-hundred-and-eleven, or 97%,
of the invited participants completed the survey. The primary objective of the
survey was to investigate the distribution of hypertension and risk factors for
hypertension in both urban and rural populations in India.

Survey questions covered demographic, anthropometric, behavioral, and dietary
characteristics, including systolic and diastolic blood pressure. Hypertension
is characterized by high blood pressure. In addition to the continuously valued
blood pressure measurements, there was also a yes/no-valued measurement of
hypertension, defined according to criteria from the World Health Organization:
either a systolic blood pressure measurement exceeding 140 mmHG or a diastolic
blood pressure exceeding 90 mmHG.

We created three separate datasets that mimiced the real blood pressure data
from the urban subpopulation studied in  @bhadoria.jfcm. These datasetes are
called `BPUrban300`, `BPUrban500`, and `BPUrban1000` and respectively contain
300, 500, and 100 hypothetical individuals from the same surveyed population.
These datasets are included in the `healthds` R package that you installed at
the beginning of this course and so they should be immediately available to you
upon loading the package by running the command `library(healthds)`.

The individuals in the `BPUrban300`, `BPUrban500`, and `BPUrban1000` datasets
are simulated, and so while the distributions of each variable and the
associations between variables are intended to mimic what one might expect to
see in this Indian population, these datasets cannot be used to make
scientific conclusions about the risk of hypertension. Rather, Dr. Mukherjee and
Dr. Boonstra will use these data in the video lectures, guided practices, and
independent guides to demonstrate the statistical ideas and R commands covered
in this course.


#### References