---
title: "The `weighted.mean` function"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In order to complete the next step of creating the indicator table, you 
will need to make use of a slightly more complicated R function: `weighted.mean`. This
document gives you an introduction to that function. 

## What is a weighted mean?

To see what a weighted mean is, think first about what a 'mean' or 'average' is. To calculate
the mean of a set of numbers, you add up the numbers and then divide that sum by the
size of your set. So, if our set of numbers if $\{0, 1, 3, 10\}$, then adding them
up gives us 0+1+3+10=14. Next, dividing 14 by the size of the set gives $14/4$, or
3.5. We would say the mean of this set of numbers if 3.5. This is sometimes called
the *simple* mean. As you've already seen, 
you can calculate the mean of a set of numbers in R using the `mean` function. 

The simple mean treats all numbers equally; the *weighted* mean puts more weight on some 
of the numbers. To calculate a weighted mean, you need to have a second set of numbers
as long as the first set; these are the weights. Suppose in our same example above 
we have weights $\{3, 2, 2, 1\}$. With these two sets, we calculate the weighted
mean as follows. First, multiply each weight by its number, so $3\times0$, $2\times1$, $2\times3$, and $1\times10$. Then, we add up those products: $3\times0+2\times1+2\times3+1\times10 = 18$. 
Then, we add up the weights themselves: $3+2+2+1=8$. Finally, divide the weighted sum
by the sum of the weights to get the weighted mean: $18/8=2.25$. Relative to the
simple mean, the number zero was given greater weight, and the number ten
was given less. For this reason, the weighted mean turns out to be smaller than 
the simple mean. 

For some people, it's helpful to think of the simple mean as being a special case of the weighted
mean when you have weights that are all equal to each other, e.g. $\{1,1,1,1\}$ or $\{2,2,2,2\}$. 

## How to calculate the weighted mean in R

As already mentioned, you'll need to use the `weighted.mean` function. We'll 
demonstrate its use in the infant growth data, but note that we will have two contexts for
the word 'weight': there is the 'weight_kg' measurement, which gives the literal
weight for each infant at each time, and there is a conceptual weight that will 
be used for the calculation of the weighted mean. 

Below I will calculate two types of means for each of the infants. The first will
be the simple mean of each infant's height over the four measurements. The second
will be a weighted mean, using the integer month (0, 1, 2, 3) as the weight. 

```{r}
library(healthds)
library(tidyverse)
growth_data %>%
  group_by(id) %>%
  summarize(mean_height_cm = mean(height_cm),
            weighted_mean_height_cm = weighted.mean(height_cm, months))
```

The important thing to see in the R code above is that we are providing *two* arguments
in the call to the `weighted.mean` function, and they are both variables taken from the `growth_data`: 
`height_cm` is the variable that we are interested in averaging over, and `months` is the variable that will provide our weights for the weighted mean.

The result is a dataframe with two rows (one for each infant) and two measurements (plus
the infant id). The first measurement is the simple mean, and they are 52.3cm and 53.3cm, 
respectively. The second measurement is the weighted mean, which are 55.4cm and 55.6cm. 
It is not too surprising that the weighted means are larger, because the infants grow
longer over time, and the weighted mean is giving more weight to height measurements
that were taken at 3 months than at 0 months. 

## Means of 1s and 0s or TRUEs and FALSEs

Sometimes you want to take the mean or weighted mean of a binary variable, say a variable that is
comprised of TRUEs and FALSEs. If the data are stored as TRUEs and FALSEs, 
R will silently coerce a TRUE to 1 and a FALSE to 0 and then calculate the mean 
of those 1s and 0s as if they were numbers. So, for example, the mean of 
$\{$TRUE, TRUE, FALSE, TRUE$\}$ will be equal to 0.75, and this is interpreted as 
the proportion of TRUEs (or 1s) in your set of numbers. This fact will be helpful
for you as you tackle the next step in constructing the indicator table. 

