---
title: "Introduction to Independent Guides"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

If you've completed Course 1 from this series, you should already be familiar 
with the purpose of independent guides for this course and do not need to read
the rest of this document. 

The pattern of learning for this course will generally proceed through the following
steps, which are the same as those used in Course 1 of this series:

1. **Learning** about ideas through lecture-style videos.
2. **Implementing** those ideas together in a guided practice video, where the
instructor demonstrates the use of specific functions in R, and learners can repeat
the steps that the instructors demonstrates.
3. Repeating Steps 1 and 2 for the all topics in a lesson. 
4. **Reinforcing** on your own by following along a series of written
steps called 'independent guides', which cover all topics in a lesson.
5. **Practicing** your understanding of these ideas through non-graded quizzes. 

In this document, we provide some context for the independent guides in Step 4, 
namely: what they will cover, when to work them, and how to use them. 

#### Independent guides will re-visit the content covered in the guided practice

For some, the most efficient way to learn to code is by watching and repeating, 
whereas for others it is by reading and copying. Step 2 (implementing via guided  
practice) relies upon the former, whereas Step 4 (reinforcing via independent guides) 
uses the latter. Thus, a defining feature of the independent guides is that what's
covered in Step 2 will be covered again in Step 4. If you don't understand 
something the instructor *says* (or even if you do understand), you will have a 
second chance to engage with the same content by reading what the instructor *writes*. 

#### Independent guides sometimes provide additional nuance or explanation

Written communication provides opportunities for nuance and multiple
simultaneous lines of thought that can be more challenging in oral
communication. In addition, topics within one lesson often share important
connections. Thus, the independent guides in Step 4, being written summations of
many videos/topics in a lesson, are able to provide additional background on a
key concept or on the relationship between multiple topics. *For this reason, we
recommend you read the independent guides for all topics.*

#### Work through the independent guides after watching the guided practice

As stated above, the independent guides are intended to complement and enhance
the guided practice videos from a single lesson. After working through all
videos in a single lesson (Steps 1 and 2), we recommend you also open the
independent guide for that lesson and then test your own understanding by taking
the practice quiz (Steps 4 and 5). In other words, working through an entire
lesson (Steps 1--5) is the most natural 'learning unit' in this course. It is
not absolutely necessary that you complete an entire lesson in one sitting, but
bear in mind that these steps are intended to be reinforcing, and so taking
breaks between lessons is preferred to taking breaks within a lesson.

#### Run R code from the independent guides in your own session

The language in the independent guides is informal and instructional. Like the
guided practice videos, these guides teach by doing, so it is assumed you will
have an R session open and that you'll be copying and pasting (or re-typing)
code from the independent guide into your R session as you read through the
guide. You will be instructed to copy and paste lines of code from the document
into your own R session. If you paste them directly into your R console and
press Enter, the code will run. Alternatively, you can paste them into
an R script, which allows you to save a copy of the steps that you run; most
analyses are saved in R scripts. However, pasting code into an R script does
not execute the code; to do so, put your cursor on the line of code you want to
run and press Ctrl+Enter (Windows) or Cmd+Enter (Mac). 

Because the independent guides will repeat everything done in the
guided practice session, you will ultimately need to decide whether you will
only actually copy and run code while watching the guided practice, only while
reading the independent guides, or both. Whatever you decide, it is recommended
that you still both watch the guided practice videos and read through
independent guides.

## Installing R packages for this course

You will need to install two R packages for this course: `tidyverse` and
`healthds`.

#### How to install `tidyverse`

To install `tidyverse`, run the following line of code in your console:

```{r, eval = FALSE}
install.packages("tidyverse")
```

Installation of `tidyverse` may take a few minutes, since your computer will
need to find the package on the internet, download it, and install it (along
with other packages that `tidyverse` requires, called dependencies). A
successful installation should print a message like this to the console:

```{r, eval = TRUE, echo = FALSE}
install.packages("tidyverse", repos = "https://cloud.r-project.org")
```

Upon successful installation, the `tidyverse` package is now installed on your 
computer but not yet loaded into your R session. Run the 
following code to load the `tidyverse` into your R session. 

```{r}
library(tidyverse)
```

Having run this, you see a couple of messages. First, because `tidyverse` is in
fact a package of packages, we are told which packages are loaded and their version numbers. 
Second, we are told that there are two functions, `filter` and `lag`, that already
existed in our R sessions. You cannot have two identically named functions at once, and
so the `tidyverse` versions are taking precedence, or 'masking', the ones that
were previously there. 

You will not need to re-install `tidyverse` if you close RStudio, but you will
need to re-load it, i.e. using the `library` command. Therefore, you should
put `library(tidyverse)` at the top of any script inside which you intend
to use functions from `tidyverse`. 

#### How to install `healthds`

Installing the `healthds` package is slightly more complicated because it has been
developed specifically for learners in this course, and it is not located in
the usual repositories that the `install.packages` function would look in. 
To install `healthds`, we first need to install and load a package called `remotes`:

```{r, eval = FALSE}
install.packages("remotes")
library(remotes)
```

```{r, echo = FALSE}
library(remotes)
```

Next, we will use a function from the `remotes` package called `install_github`. 
This function acts similarly to `install.packages` but allows us to install
packages that are hosted on GitHub, a repository where any user can develop and
publicly share software. To install `healthds`, run the following line 
of code (it is easiest to copy-paste this directly into your R console):

```{r, eval = FALSE}
install_github("umich-biostatistics/healthds", upgrade = "never")
```

If you run the above line of code without first having installed and loaded
`remotes`, you will get an error message that the function `install_gitub` is not 
found. Just as with `tidyverse`, you must load `healthds` using the `library`
command. 

To summarize, for any independent guide that you do in this course, 
you should put the following commands at the top of your script to ensure that
you have access to the functions that they contain:

```{r}
library(tidyverse)
library(healthds)
```

And execute these commands by putting your cursor on each line and pressing
Ctrl+Enter (Windows) or Cmd+Enter (Mac).



