---
title: "Module 1, Lesson 2: Tests for comparing two proportions"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

Welcome to the independent guide for the first lesson of Module 1. 

In this course, we will use the `tidyverse` and `healthds` R packages. If you
have not yet installed these on your computer, you should do so now (you only
need to install them once). For help on installing these packages, please refer
to the instructions provided at the beginning of this course.

Now, open a new R script. At the top of this script, write the following:

```{r, message = F, warning = F}
library(healthds)
library(tidyverse)
```

You can copy this R code by pressing the small symbol in the top-right of each
box and then pasting it into your own script. However, the script just putting
this code into your script does not execute the code. To run a line of code from
your script, put your cursor on the line of code to be run and press
'Ctrl+Enter' (on your Windows computer) or 'Cmd+Enter' (on your Mac).

### Comparing proportions across two groups

In this independent guide, we will learn about two ways to compare proportions
between two groups. The first is the two-sample test of proportions, as
implemented in the R function `prop.test`, and the second is Fisher's exact
test, as implemented in the `fisher.test` function.

To demonstrate these functions, we will analyze the `BPUrban1000` data, which
comprises health information on 1000 hypothetical individuals from India. In
particular, we want to know whether the proportion of individuals who report
having hypertension is different among those who are smokers versus those who
are not smokers.


#### Preparatory data wrangling

To prepare the data for analysis, we want to create a two-by-two table, where
the rows of the table indicate the person's smoking status based upon the
`smoking` variable (`smoking = FALSE` corresponds to non-smokers, `smoking =
TRUE` corresponds to smokers). It is this two-by-two table that is required
by the `prop.test` and `fisher.test` functions. 

Here is some code that *almost* gets us the two-by-two table that we need, using
ideas that we covered in Course 1:

```{r}
BPUrban1000 %>%
  group_by(smoking) %>%
  summarize(num_htn = sum(hypertension),
            num_no_htn = sum(1 - hypertension))
```

Walking through this code step by step, we first take the `BPUrban1000` data and
group it by the `smoking` variable. As a consequence, subsequent actions will
be conducted separately for each unique category of `smoking` (in this case `FALSE`
and `TRUE`). There are two specific actions we take: we calculate the
number of individuals with hypertension (`sum(hypertension)`) and save it as
a variable called `num_htn`, and we calculate the number of individuals without
hypertension (`sum(1 - hypertension)`) and save it as a variable called
`num_no_htn`. 

The result is a dataset with two rows (one for each smoking
group) and three columns: the first column indicates the smoking category
and the second and third columns give the counts of those with hypertension
and without hypertension, respectively. 

If you run the line of code above, the output will only be printed to the screen
and not saved as an object to be used later, e.g. passed into the `prop.test`
and `fisher.test` functions. We want to use the assignment operator (`<-`) to
save this as object, and we'll name our new object `smoking_by_hypertension` (as
seen below).

In addition, we need to make some minor adjustments to the result in order to
make it into an acceptable form. First, we want the first column not to be a
literal column but rather an "implicit" column that denotes the label of each
row (just like we have an implicit row that denotes the name of each column). We
use the `column_to_rownames` function to do so. Second, the `prop.test` and
`fisher.test` functions require that we provide a numeric `matrix` rather than a
`data.frame`. Formally discussing the differences between a `matrix` and
`data.frame` is beyond the scope of this course.

Taking into account these extra steps, the code below implements everything that
is needed to translate our data into the appropriate two-by-two table for
testing whether there is a difference in hypertension proportions between
smoking groups:

```{r}
smoking_by_hypertension <- 
  BPUrban1000 %>%
  group_by(smoking) %>%
  summarize(num_htn = sum(hypertension),
            num_no_htn = sum(1 - hypertension)) %>%
  column_to_rownames("smoking") %>%
  as.matrix()
```

Copy and paste this code into your script, and then run it. Having created this
object, we can also see what it looks like by typing its name:

```{r}
smoking_by_hypertension
```

This is the two-by-two table that we need. The two rows indicate non-smokers
and smokers, respectively, the two columns indicate hypertension and no 
hypertension, respectively, and the numbers inside indicate the counts of the
number of people with each combination. 

#### Using `prop.test` to compare two proportions

The two-sample test of proportions is implemented in `prop.test`. After creating
the `smoking_by_hypertension` above, you can plug that directly into the
function:

```{r}
prop.test(x = smoking_by_hypertension)
```

Having run this code, you should see some results printed to the screen. At the
top of the results, we see the test that was conducted "2-sample test for
equality of proportions with continuity correction". 

Jumping down to the very bottom, we see the two observed proportions that are
being statistically compared. The proportion of non-smokers who have
hypertension is 0.2024, which is 186 (the number of non-smokers with
hypertension) divided by 186 + 733 (the total number of non-smokers). Similarly,
the proportion of smokers who have hypertension is 0.3210, which is 26 (the
number of smokers with hypertension) divided by 26 + 55 (the total number of
smokers). Interestingly, it does not print out the value for the difference in
these proportions, but with some arithmetic we can determine that this
difference is 0.2024 minus 0.3210, or -0.1186. And so we would estimate that the
true proportion of hypertension in non-smokers is about 12% lower than in smokers.

Although we don't see the estimated difference in this output, we do see an
estimated 95% confidence interval for this difference. The lower bound of this
confidence interval is -0.2302, and the upper bound is -0.0070.

Looking up a little higher, we see that the p-value corresponding to this test
is 0.0182. Were we to assume that in truth there is no difference in
the true proportion of hypertension between smokers and non-smokers, this p-value is
the probability of observing our data or data that are more extrem", where 
"more extreme" means  a difference that is even farther away from zero than -0.1186. Since the
p-value is small, we might conclude that it is unlikely that our data arose from
a scenario in which the true proportions are equal.

There is an alternative way to use `prop.test` by typing in the numbers in 
each group with hypertension as the value for the `x` argument and the total 
numbers in each group as the value for the `n` argument:

```{r}
prop.test(x = c(186, 26), n = c(186 + 733, 26 + 55))
```

This will yield the same statistical results as the first way. 


#### Using `fisher.test` to compare two proportions

The `fisher.test` implements Fisher's exact test:

```{r}
fisher.test(x = smoking_by_hypertension)
```

We are testing whether the 'odds ratio' comparing hypertension in non-smokers
against smokers is equal to 1. Although the odds ratio is not the same as a
difference in proportions, we can say that an odds ratio of 1 (which is
the null hypothesis) corresponds to there being no difference in proportions

How do odds ratios and proportions relate? The *odds ratio* is, as its name
suggests, a ratio of two odds, and an *odds* is a proportion divided by one
minus a proportion. From our work above we know that the observed proportion of
non-smokers with hypertension is 0.2024, meaning that the odds of hypertension
among non-smokers is 0.2024 / (1 - 0.2024), or about 0.2538. Similarly, the odds
of hypertension among smokers is 0.3210 / (1 - 0.3210), or 0.4726. Thus, the
odds ratio is 0.2538 / 0.4726, or 0.5370, which is very close^[In statistics,
there are generally many different approaches to estimating parameters such as
odds ratios, and `fisher.test` uses a slightly different estimation approach
than what we have done, which is simply calculating the 'sample odds ratios'.
Thus, the results are similar but not identical] to the output from
`fisher.test`. From this, we would conclude that the odds of hypertension for
non-smokers are nearly half as large as the odds of hypertension for smokers.

We are also provided a 95% confidence interval for this odds ratio, the lower
bound of which is 0.3211 and the upper bound of which is 0.9180. Because the
entire interval falls below 1, we would conclude that the odds of hypertension 
among non-smokers seem to be less than the odds of hypertension among smokers. 
Qualitatively, this is the same conclusion as that of the two-sample test of
proportions, namely that the non-smokers have lower rates of hypertension than
smokers. 

What would happen to our statistical conclusions if we were to switch the order 
of the rows in the underlying data before conducting the test? Switching the 
order of the rows can be done using the command `smoking_by_hypertension[c(2, 1), ]`.
Running Fisher's exact test on these modified data is done as follows:

```{r}
fisher.test(smoking_by_hypertension[c(2, 1), ])
```

At first the results look much different, but upon closer inspection, we can see
that the conclusions remain the same. The roles of the two smoking groups (and
the numerator and denominator of the odds ratio) have switched. Now the odds
ratio is estimated to be 1.8615, and this is interpreted to mean that the odds
of hypertension among smokers is almost twice as large as the odds of
hypertension among non-smokers. The confidence interval is for this alternative
odds ratio, and now it falls entirely above 1. Finally, the p-value is
identically equal to our first result. If we were to invert this odds ratio, we
would see that 1 / 1.8615 equals 0.5372, which is the original estimated odds
ratio. Thus, although we must interpret the results differently, our conclusion
remains the same.








