---
title: "Introduction to Independent Guides"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---


```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

The pattern of learning for this course will generally proceed through the following
steps:

1. **Learning** about ideas through lecture-style videos.
2. **Implementing** those ideas together in a guided practice video, where the
instructor demonstrates the use of specific functions in R, and learners can repeat
the steps that the instructors demonstrates.
3. Repeating Steps 1 and 2 for the all topics in a lesson. 
4. **Reinforcing** on your own by following along a series of written
steps called 'independent guides', which cover all topics in a lesson.
5. **Practicing** your understanding of these ideas through non-graded quizzes. 

In this document, we provide some context for the independent guides in Step 4, 
namely: what they will cover, when to work them, and how to use them. 

#### Independent guides will re-visit all content covered in the guided practice

For some, the most efficient way to learn to code is by watching and repeating,
whereas for others it is by reading and copying. Step 2 (implementing via guided 
practice) relies upon the former, whereas Step 4 (reinforcing via independent
guides) uses the latter. Thus, a defining feature of the independent guides is that
everything covered in Step 2 will be covered again in Step 4. If you don't understand
something the instructor *says* (or even if you do understand), you will have a second chance 
to engage with the same content by reading what the instructor *writes*. 

#### Independent guides sometimes provide additional nuance or explanation

Written communication provides opportunities for nuance and multiple simultaneous lines of 
thought that can be more challenging in oral communication. In addition, topics
within one lesson often share important connections. Thus, the independent guides 
in Step 4, being written summations of many videos/topics in a lesson, are able to provide
additional background on a key concept or on the relationship between multiple 
topics. *For this reason, we recommend you read the independent guides for
all topics.*

#### Work through the independent guides after watching the guided practice

As stated above, the independent guides are intended to complement and enhance
the guided practice videos from a single lesson. After working through all videos in a single lesson 
(Steps 1 and 2), we recommend you also open the independent guide for that lesson and
then test your own understanding by taking the practice quiz (Steps 4 and 5). 
In other words, working through an entire lesson (Steps 1--5) is the most natural 
'learning unit' in this course. It is not absolutely necessary that you complete 
an entire lesson in one sitting, but bear in mind that these steps are intended to be 
reinforcing, and so taking breaks between lessons is preferred to taking
breaks within a lesson. 

#### Run R code from the independent guides in your own session

The language in the independent guides is informal and instructional. Like the
guided practice videos, these guides teach by doing, so it is assumed you will have an R 
session open and that you'll be copying and pasting (or re-typing) code from the 
independent guide into your R session as you read through the guide. You will be
instructed to copy and paste lines of code from the document into your own R session.

However, because the independent guides will repeat everything done in the guided 
practice session, you will ultimately need to decide whether you will only actually copy 
and run code while watching the guided practice, only while reading the independent 
guides, or both. Whatever you decide, it is recommended that you still both 
watch the guided practice videos and read through indepedent guides. 

