---
title: "Independent Guide: 2.3 - dplyr"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the third lesson of Module 2, you were introduced to additional functions from
the dplyr package. To practice using these functions, open the R project called
`Lessons.Rproj` and a blank R script. Before we can begin, you'll also need to
load the necessary packages into the current R session.

```{r}
library(healthds)
library(tidyverse)
```

## `pivot_wider()`

Replicating the example from the lecture, we can pivot the `growth_data` from
long format into wide format using the `pivot_wider` function so that all
measurements from an infant are contained in a single row.

To use the `pivot_wider` function, there are a few required arguments as well
as several optional arguments. In general, as long as the arguments are
explicitly named, the arguments do not have to be provided in any particular order. 
The `id_cols` argument specifies the variable that 
uniquely identifies each row, which is the infant ID variable in this example. 
The `names_from` argument
specifies the variable that indicates the repeated measurements within the
subject ID (e.g. the months variable). Lastly, the `values_from` argument
specifies the variable to use for filling in the cell values of the table (e.g.,
the weight variable).

```{r}
growth_data %>%
  pivot_wider(id_cols = id,
              names_from = months,
              values_from = weight_kg)
```

R doesn't allow numeric variable names and thus, in the printed result, it 
has put so-called 'backticks' around the number in each column to create valid 
variable names. However, these variable names are not particularly informative 
of the data contained in each
column. In this case, the `names_prefix` argument is useful for adding a string
before every variable name. Returning to our example, this argument allows you
provide a prefix that conveys both the timing of the measurements (i.e. month)
and the type of measurement (i.e. weight):

```{r}
growth_data %>%
  pivot_wider(id_cols = id,
              names_from = months,
              names_prefix = "weight_month",
              values_from = weight_kg)
```

Let's assign this data frame to an object:

```{r}
growth_wide_weight <-
  growth_data %>%
  pivot_wider(id_cols = id,
              names_from = months,
              names_prefix = "weight_month",
              values_from = weight_kg)
```

Recall that the `pivot_wider` and `pivot_longer` functions are inverses of each
other. The `pivot_longer` requires the `cols` argument to identify the columns
to pivot into long format, and we can also use the NOT operator (`!`) to denote
"all columns except..." within this argument. Below we pivot all columns except the
infant ID variable:

```{r}
growth_wide_weight %>%
  pivot_longer(cols = !id)
```

If we want to change the variable names to something more informative, we can
use the `names_to` and `values_to` arguments:

```{r}
growth_wide_weight %>%
  pivot_longer(cols = !id,
               names_to = "months",
               values_to = "weight_kg")
```

Notice that the results resemble the original `growth_data`, demonstrating that
the `pivot_wider` anf `pivot_longer` functions are natural complements of each
other.

Remember to save the work you've done. Try saving the script you've been working
in as `pivot_wider.R` for later reference.

## `mutate()`

To practice the `mutate` function from the dplyr package, open another blank
script within the same R project.

Let's consider the task of creating a BMI variable using the height and weight
measurements from the `growth_data`. Within the `mutate` function, we name the
new variable to create and specify how this variable is defined (i.e. BMI
formula).

```{r}
growth_data %>%
  mutate(bmi = weight_kg / (height_cm / 100)^2)
```

Notice that this returns the original `growth_data` along with the new variable
that has performed the requested calculations.

The nice thing about the `mutate` function is that we can create multiple new
variables within one function call and even reference variables that were just
previously created in that same function call. For example, we can create
a height variable that performs a unit conversion from centimeters to meters
prior to creating the BMI variable.

```{r}
growth_data %>%
  mutate(height_m = height_cm / 100,
         bmi = weight_kg / height_m^2)
```

This code returns the original `growth_data` along with two new columns, where
the calculated values for BMI are the same as before.

We could assign this data frame to an object, particularly if we plan on
performing subsequent calculations on the BMI measurements. Without assignment,
R does not save the new variables that were created.

```{r}
growth_data_expanded <-
  growth_data %>%
  mutate(height_m = height_cm / 100,
         bmi = weight_kg / height_m^2)
```

Just as before, remember to save the work you've done. Try saving the script
you've been working in as `mutate.R` for later reference.

## `arrange()` and `select()`

Some of your data-wrangling steps will not be for purposes of changing the information in the data but rather focusing your attention on certain columns using `select` or sorting rows using `arrange`. To practice the `arrange` and `select` functions from the dplyr package, open another blank script within the same R project.

The `arrange` function requires the variable by which to sort or arrange the
data. The following code sorts the data by the month variable in ascending
order:

```{r}
growth_data %>%
  arrange(months)
```

We can also sort the data by multiple variables. In this case, the data is first
sorted by the month variable followed by the weight variable within tied values
of the month variable:

```{r}
growth_data %>%
  arrange(months, weight_kg)
```

The default setting is for variables to be sorted in ascending order, but
variables can also be sorted in descending order using the `desc` function:

```{r}
growth_data %>%
  arrange(desc(months), weight_kg)
```

To be clear, what the above code does is sort first by `months` in *decreasing* order and then, among observations having common values of the motnh, sorting by increasing (the default direction) values of `weight_kg`. 

The `select` function chooses columns in the data frame:

```{r}
growth_data %>%
  select(id, months)
```

Using the NOT (!) operator, we can also select everything except a particular
column:

```{r}
growth_data %>%
  select(!weight_kg)
```

The `select` function is particularly useful for re-ordering columns:

```{r}
growth_data %>%
  select(id, weight_kg, height_cm, months)
```

However, it's not always easy to name every column, especially if there are
many columns in your data frame. In this case, there are helper functions that
help us more easily select columns. For example, we can use the `everything`
function to name columns prior to selecting the rest of the columns.

```{r}
growth_data %>%
  select(months, everything())
```

The `starts_with` function allows us to provide the beginning of the column name
that we would like to select.

```{r}
growth_data %>%
  select(months, starts_with("w"))
```

Similarly, the `ends_with` function allows us to provide the end of the column
name that we would like to select.

```{r}
growth_data %>%
  select(months, ends_with("cm"))
```

The `contains` function matches the string argument anywhere in the column name.
For example, the "ght" string returns both the height and weight variables to
select.

```{r}
growth_data %>%
  select(months, contains("ght"))
```

Remember to save the work you've done. Try saving the script you've been working
in as `arrange_select.R` for later reference.
