---
title: "Module 3, Lesson 3: Advanced topics in ggplot"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the third lesson of Module 3, you delved into some more advanced topics
in ggplot. Layering geoms can tell multiple stories about your data, and we saw
that you can create fairly complicated plots with simple code by leveraging 
concepts such as 'positional matching' (a concept you can use for *any* R function, 
not just `ggplot2` functions) and 'inheritance', which is a 'ggplot2'-specific feature. 

You learned how to begin to fine-tune your plots using the `scale` family of
functions. And finally, you learned how to make use of the concept of small multiples
to create faceted plots. 

Open the R project called `Lessons.Rproj` and a blank R script. Then, 
load the necessary packages into the current R session:

```{r}
library(healthds)
library(tidyverse)
```


## Layering, positional matching, and inheritance

The standard `ggplot` template is not too onerous to write when creating a single
geometric object. For example, a boxplot showing the distribution of weight
measurements for each infant is created by

```{r}
ggplot(data = growth_data) + 
  geom_boxplot(mapping = aes(x = weight_kg, y = id))
```

However, when showing boxplots, it is usually good practice to also show the underlying
data, especially when your data are small-ish. To do so, we layer on a call
to `geom_point`:

```{r}
ggplot(data = growth_data) + 
  geom_boxplot(mapping = aes(x = weight_kg, y = id)) +
  geom_point(mapping = aes(x = weight_kg, y = id))
```

The code starts to get a bit more complicated looking, and we can trim things down
to make it cleaner. First, we can lean on three key facts: (i) the first
argument of `ggplot` is `data`; (ii) the first argument of every `geom_<GEOMNAME>`
is `mapping`; (iii) and the first and second arguments of `aes` are `x` and `y`, respectively. 

Because of these key facts, we can use the R concept of 'positional matching', which
allows you to specify arguments in their expected order *without naming them*. So, 
the code below is equivalent to the above code:

```{r, eval = FALSE}
ggplot(growth_data) + 
  geom_boxplot(aes(weight_kg, id)) +
  geom_point(aes(weight_kg, id))
```

In particular, we don't need `data = ` in the call to `ggplot` because the `data`
argument comes first, and therefore `growth_data`, which is now an 
'unnamed' argument, is matched to `data` by virtue of its position. The same
principle applies to justify not needing to write `mapping = `, `x = `, and `y = `.

There's an additional way we can make this code even more concise using 'inheritance', 
which is a `ggplot` specific feature. Inheritance says that we can place inside of 
`ggplot` all elements of a plot that are intended to be shared in common by every layer. 
Referring to our figure, one shared element is `data`, and inheritance is already
happening for this: the provided `data` object is silently passed down to and
used by `geom_boxplot` and `geom_point`. Another shared element are the mapped
aesthetics, as specified by `aes(weight_kg, id)`. For this reason, the `ggplot`
function has its own mapping argument (which is positioned second in the 
set of arguments), which, when specified, also gets passed down to the geoms. 

Thus, the following code is also equivalent to all of the above code chunks:

```{r, eval = FALSE}
ggplot(growth_data, aes(weight_kg, id)) + 
  geom_boxplot() +
  geom_point()
```

Now, both the `data` argument (the first argument in our call to `ggplot`, 
by virtue of positional matching) and the `mapping` argument (the second argument,
for the same reasons) get passed down to each of the geoms. 

Inheritance is very flexible. If there's an aesthetic mapping that you want to
apply to one geom, you can still incorporate it in the usual way without 
affeting the inheritance:

```{r, eval = FALSE}
ggplot(growth_data, aes(weight_kg, id)) + 
  geom_boxplot() +
  geom_label(aes(label = months))
```

In the above plot, the points are labeled according the value of `months`. You 
can also overwrite an aesthetic that would otherwise be inherited from `ggplot`. 

Before continuing, save your work thus far. Save your script as `03_03advancedggplot.R`.

## The scale family of functions

Please note that in the remaining sections of this independent guide, we will 
still cover all of the  same topics discussed in the corresponding guideded 
practice but with slightly different examples.

The code below creates several plots. We first create a plot showing individual
pairs of height-weight measurements, using different colors for each infant id. 
We then layer on the infant-specific height-weight trajectories. We have
done so while taking full advantage of positional matching and inheritance, but
we also use a novel feature that we've previously not used:

```{r}
weight_versus_height <- 
  ggplot(growth_data, aes(weight_kg, height_cm, color = id)) + 
  geom_point()

weight_versus_height
```

We've first *assigned* the output from ggplot to an object named `weight_versus_height`. 
The plot isn't shown when we make the assignment, but if we type the name of the 
object in the console and press  enter, we view the plot. It is the 'base' plot 
showing just the points.

```{r}
weight_versus_height +
  geom_line(aes(color = id))
```

With this plot object created, we can extend it by adding on other `ggplot` 
features before printing, as you can see in the chunk above. Specifically, 
we've layered on a call to `geom_line` to show the trajectories.

To make the variable labels more informative, we use the `scale_<AESTHETIC>_continous`
and `scale_<AESTHETIC>_discrete` families of functions. Every aesthetic has its
own set of scale functions. Within these functions, we can modify the way in which 
the aesthetic is visualized and presented. So, for example, to change the labels, we 
would write the following:

```{r}
weight_versus_height + 
  geom_line() +
  scale_y_continuous(name = "Height (cm)") +
  scale_x_continuous(name = "Weight (kg)") +
  scale_color_discrete(name = "ID")
```

These scale features are added to the baseplot just like other functions in `ggplot2`.
We have to use the *continuous* versions of these scale functions for the `x`
and `y` aesthetics because they are doubles, whereas we have to use the *discrete*
version for the `color` aesthetic, because the `id` variable is a character. 

The quoted names can be whatever you choose. Take care in writing out these aesthetic names. 
If you accidentally switch the values of the `name` argument for the `x` and `y` 
aesthetics, the labels of the x-axis the y-axis will be switched, and the
plot will be misleading. 

The scale family of functions allows for other useful modifications of your plot. 
A common example is to change the number or nature of the tick marks along an axis. 

```{r}
weight_versus_height +
  geom_line() +
  scale_y_continuous(name = "Height (cm)", 
                     breaks = seq(from = 48, to = 58, by = 2)) +
  scale_x_continuous(name = "Weight (kg)") +
  scale_color_discrete(name = "ID")
```

Here I am using a function from base R called `seq`, which is used to create
a sequence of numbers stored as a vector. In this case, I'm creating a vector
that starts at 48, increments by 2 at a time, and continues on in this manner
until it would go beyond a number larger than 58, stopping just before this point.

You can alternatively use the concatenate function, `c()`, to provide a
custom set of breaks, as in:

```{r}
weight_versus_height +
  geom_line() +
  scale_y_continuous(name = "Height (cm)", 
                     breaks = c(50, 51, 53, 57)) +
  scale_x_continuous(name = "Weight (kg)") +
  scale_color_discrete(name = "ID")
```

To see what other features of an aesthetic you can modify with the `scale` family
of functions, you should inspect the help page, e.g. `?scale_x_continuous`. 

If you only need to change the labels of aesthetics and nothing else, there is
a shortcut function provided called `labs`:

```{r}
weight_versus_height +
  geom_line() +
  labs(y = "Height (cm)", 
       x = "Weight (kg)",
       color = "ID")
```

Save your script at this point using `Cmd+S` or `Ctrl+S` and then keep
reading to add more code to this script. 

## `facet_grid` and `facet_wrap`

Faceting uses the idea of small multiples to show data across time or space. Faceting
is sort of like another aesthetic, allowing you to show your data across multiple
panels. To highlight how to facet as well as to demonstrate that we can add 
any `ggplot` function to an existing plot (and not just `scale` or `geom_` 
functions as above), run the following code:

```{r}
weight_versus_height + 
  facet_grid(rows = vars(months))
```

As an aside: this code will only run if you created the `weight_versus_height` object
as we did above. Otherwise, you will get an error message from R that states

```{r, echo = FALSE}
cat("Error: object 'weight_versus_height' not found")
```

Here we have taken our base plot of weight versus height, which we created
earlier, and added instructions to plot each id separately. The `vars` function
acts similarly to the `aes` function and should contain the name or names of 
variables that we wish to facet by, separated by commas. The `rows` argument 
in `facet_grid` means that each unique combination of the variables in `vars` 
will get its own row. To facet by columns, we would use the following:

```{r}
weight_versus_height + 
  facet_grid(cols = vars(months))
```

Save your script once again before closing.