---
title: "Module 3, Lesson 1: The basics of the grammar of graphics"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In this first lesson of Module 3, you were introduced to the `ggplot2` package, 
which allows you to create publication-quality graphics with relative ease. 
The package is part of the `tidyverse` and so is already installed and available
to you on your computer. Open the R project called
`Lessons.Rproj` and a blank R script. As always, you'll first need to
load the necessary packages into the current R session:

```{r}
library(healthds)
library(tidyverse)
```

## The `ggplot` template

Here is the basic template for creating plots in ggplot:

```{r, echo = TRUE, eval = FALSE}
ggplot(data = <DATA>) + 
  geom_<GEOMNAME>(mapping = aes(<MAPPINGS>))
```

It is only a *template*, and it won't run correctly until you replace the 
bracketed expressions, namely `<DATA>`, `<GEOMNAME>`, and `<MAPPINGS>`, with
appropriate R code. 

The `<DATA>` expression should be replaced with the name of a dataset that
exists in your R session. In our guided practice, we exclusively use `growth_data`. 

The `<GEOMNAME>` expression should be replaced with any valid geometric (geom)
object. Perhaps most common is `geom_point`, which creates a scatterplot, i.e.
plots individual points. Many other choices exist too, some of which we'll learn 
about in this module. 

The `<MAPPINGS>` expression should be replaced by one or more equations, separated
by commas, that *map* certain variables in your dataset to certain aesthetics. 
The `x` and `y` aesthetics, respectively controlling the x-axis and y-axis, 
are the most common aesthetics, and there are several others. 

Note also that when using multiple `ggplot2`-specific functions, you *add* 
the functions together using the `+` sign (not the `%>%` operator as you might 
expect). 

Putting this all together, we can create a simple scatterplot of the infant
growth data, plotting the height measurements on the y-axis against the weight
measurements on the x-axis:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm))
```

## Using aesthetics in `geom_point`

In `ggplot2`, aesthetics are visual cues, which may either be directional, like
the `x` and `y` aesthetics, or shape-, color-, or size-based. To see this more concretely, 
we can map the `id` variable in `growth_data` to the `shape` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, shape = id))
```

Notice that a legend corresponding to the shape aesethetic is automatically
added to the right of the figure. Although it may not be evident at first glance,
the legend is analogous to the x-axis and y-axis: just like the axes are 
instructions for interpreting values of the `x` and `y` aesthetics, legends
are instructions for interpreting the other aesthetics. 

We could alternatively map the `id` variable in `growth_data` to the `color` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, color = id))
```

You can map any variable in your dataset to (nearly) any aesthetic, but not all
choices will be equally desirable. As an example, consider the plot that maps
the `months` variable to the `size` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, size = months))
```

This plot is strange-looking because ggplot is interpreting `months` as a number --
in particular a number with small and large values -- and so months with "small"
values get a smaller size, and months with "large" values get a larger size. So, 
while it technically allows us to distinguish between different values of 
`months`, it is also somewhat distracting and potentially even misleading. 

The same basic conclusion would apply if we were to map `months` to the `alpha`
aesthetic, which controls the transparency of each point:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, alpha = months))
```

or if we were to map `months` to the `color` aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, color = months))
```

Notice that the legend for color is a so-called color-bar, in which there are 
intermediate shades reserved for fractional values of `months`. Since we do not
have fractional values of `months` in our data, this is probably not the best
way to distinguish between months. 

You may be wondering why -- when we mapped `id` to the `color` aesthetic a few plots
above -- the legend had two completely different colors, whereas when we mapped `months`
to the `color` aesthetic, a colorbar was used. The difference is that `id` is
considered to be a discrete-valued variable, whereas `months` is considered
to be a number. You can see this by printing `growth_data` to the screen:

```{r}
growth_data
```

Just below each column name is a bracketed expression. Under `id` is `<chr>`, 
indicating that this is stored as a *character*, which is a flexible data-type
that can be basically any collection of letters and numbers. In contrast, under
`months` is `<dbl>`, indicating that this is stored as a *double*, a computer-ish
term for a number. Characters are more flexible than doubles, meaning that
one can always turn a double into a character, but not necessarily the other way
around. And if `ggplot` encounters a character, it will treat it like a discrete-
valued variable. 

What this means is that, if we wanted to tell `ggplot` to treat `months` like a *character* instead of a *double*, we could run the following code:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm, color = as.character(months)))
```

Now, we get a discrete-valued legend for months, with one distinctive color for 
each unique value. We have used the `as.character()` function to coerce the `months`
variable into a character. 

Save this script as `03_01grammarofgraphics.R`. 

## Practicing new geoms

In this section you will learn about other geoms besides `geom_point`. Let's
consider two geoms that replace the points with labels that can be treated as
a new aesthetic:

```{r}
ggplot(data = growth_data) + 
  geom_text(mapping = aes(x = weight_kg, y = height_cm, label = id))
```

```{r}
ggplot(data = growth_data) + 
  geom_label(mapping = aes(x = weight_kg, y = height_cm, label = id))
```

Both `geom_text` and `geom_label` require that you map an additional aesthetic, 
called `label`. Here, we have mapped the `id` variable to the `label` aesthetic. 
In place of each point is either `001` or `002`, which are the values taken directly
from the `id` variable in `growth_data`. Compare the two plots to see that 
`geom_text` provides the "naked" values whereas `geom_label` frames each value
in a contrasting box. 

A lineplot is constructed through the use of `geom_line`:

```{r}
ggplot(data = growth_data) + 
  geom_line(mapping = aes(x = weight_kg, y = height_cm))
```

Pay close attention to how the line segments are connected. Pairs of points are
defined by neighboring values of the variable mapped to the `x` aesthetic and
*not* the `y` aesthetic. Thus, `geom_line` may sometimes result in a line that
squiggles up and then down because the `y` values are not guaranteed to be increasing,
but it will never squiggle left and then right because the `x` values are explicitly
paired to be increasing. 

Finally, you can construct a column plot using `geom_col`:

```{r}
ggplot(data = growth_data) + 
  geom_col(mapping = aes(x = weight_kg, y = height_cm))
```

As we've already noted, this plot should be interpreted with care. `geom_col`
creates columns that point up starting at zero: the horizontal position of
each column is determined by the `x` value and the vertical height is equal to 
the `y` value. In this case, there are two observations having `x=5.5`, and so
because they cannot occupy the same space, `geom_col` stacks the resulting columns. 
However, it is not obvious from simply looking at the plot that this is what has
happened. 

Save your script to preserve the additional work you've done. 

## Layering geoms

The `ggplot2` package is extremely customize-able, and one of the ways in which 
this manifests is in its ability to layer multiple geoms in one figure. To 
start, let us first create a line plot showing the relationship between height
and weight, mapping the `id` variable to the color `aesthetic`:

```{r}
ggplot(data = growth_data) + 
  geom_line(mapping = aes(x = weight_kg, y = height_cm, color = id))
```

From our experience working with `growth_data`, we know that each line is comprised
of four points, but it is basically impossible to see that here. On the other hand, 
this line plot does an exceptional job of showing the relationship between weight and height. 
We can achieve both of these, i.e. showing the individual points *and* showing
the trajectories, by layering. To layer, you add -- literally -- calls to multiple geoms. 
Here I am layering on a call to `geom_point`:

```{r}
ggplot(data = growth_data) + 
  geom_line(mapping = aes(x = weight_kg, y = height_cm, color = id)) +
  geom_point(mapping = aes(x = weight_kg, y = height_cm))
```

Finally, note that the ordering in which the layers are written matters. Visually
compare the plot above, in which `geom_line` comes first and `geom_point` comes
second, to the plot below, in which the order is reversed:

```{r}
ggplot(data = growth_data) + 
  geom_point(mapping = aes(x = weight_kg, y = height_cm)) +
  geom_line(mapping = aes(x = weight_kg, y = height_cm, color = id))
```

In the first plot, the points cover a small portion of the line because they 
are in the second layer, whereas in the second plot the lines cover a small
portion of the points. In this example, it does not matter too much
which plot is used; however, this will not always be the case. In general, 
it is good to think carefully about the order in which your geoms are layered 
and whether there is an order that makes more sense. 

As before, remember to save your work in this script. 
