---
title: "Module 2, Lesson 2: `dplyr`"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the second lesson of Module 2, you were introduced to `dplyr`, which is a package
included in the `tidyverse` that provides a set of verbs for data wrangling.

To practice using the functions from the `dplyr` package, open the R project called
`Lessons.Rproj` and a blank R script. On the first two lines of this and every 
other script we create in these Independent Guides, put the following lines of code to 
load `healthds` and the `tidyverse`.

```{r}
library(healthds)
library(tidyverse)
```

Run these lines of code the 'usual' way ('Cmd+Enter' or 'Ctrl+Enter'). 

## `filter()`

The `filter` function allows us to include or exclude observations
(i.e., rows) from a dataset based on one or more conditions. The general
structure for using this function follows: `filter(data, condition)`.

Let's consider the `growth_data`, which is now available to use via `healthds`. 
First, we can take a look at the dataset before any filtering is performed. 
There are eight observations across four variables:

```{r}
growth_data
```

If you want to filter out the infants' three-month observations, you would write 
the following line of code:

```{r}
filter(growth_data, months == 3)
```

Notice the use of *two* equal signs in the above line of code. In R, two equal
signs is used to denote "is equal to", which is distinct from using a single
equal sign to set an argument value. Think of two equal signs as asking a question 
and one equal sign as making a declaration. 

Similarly, you would write the following line of code if you wanted to filter
out the infants' two-month observations:

```{r}
filter(growth_data, months == 2)
```

The `filter` function can also be used to specify conditions other than
equality. For example, we can filter by a two-sided inequality by using `!=` to
denote "is not equal to". It is possible to filter by one-sided
inequalities as well, which are represented in R as `<`, `<=`, `>`, and `>=`:

```{r}
filter(growth_data, months != 2)
filter(growth_data, months > 2)
```

We can also filter by multiple conditions using the AND (`&`, an ampersand) and OR (`|`, 
a vertical bar) operators. The AND operator returns the observations where both conditions are
true, and the OR operator returns the observations where at least one condition
is true:

```{r}
filter(growth_data, weight_kg  > 5.0 & height_cm > 50)
filter(growth_data, weight_kg  > 5.0 | height_cm > 50)
```

One note: you can filter by multiple conditions separated by commas:

```{r}
filter(growth_data, weight_kg  > 5.0, height_cm > 50)
```

Comma-separated conditions are interpreted as AND operators, so you are free to
use either commas or ampersands. 

Remember to save the work you've done. Save the script you've been working
in as `filter.R` for later reference.

## `group_by()` and `summarize()`

To practice additional functions from the dplyr package, open another blank
script. 

The `group_by` function allows us to group data into sub-datasets defined by a
common value. In turn, the `summarize` function provides a summary of the data
based on the grouped data, where a summary statistic is returned for each
sub-dataset. Therefore, the `group_by` function effectively splits the data, and
future actions, such as the `summarize` function, are applied separately to each
sub-dataset. The general structure for using both functions follow:
`group_by(data, variable)` and `summarize(grouped_data, <SUMMARY_FUNCTION>(variable))`, 
where you replace `<SUMMARY_FUNCTION>` with an actual function name. 

For example, if we want to assess the typical weight of infants
using the `growth_data`, it would make sense to group the observations by month
and obtain the average, i.e. mean, weight for each month rather than combine all
observations across time into a single average.

First, we would start by grouping the data by month:

```{r}
group_by(growth_data, months)
```

This returns a grouped dataset that isn't does look too different
from the original `growth_data`.

Next, we can wrap this command in a summarize command by specifying the grouped
dataset as the first argument of the `summarize` function. The second
argument will be the desired summary function, which in this case is the mean
of the weight variable:

```{r}
summarize(group_by(growth_data, months), mean(weight_kg))
```

This command returns a dataset with number of rows equal to the number of
sub-datasets. The first column is defined by the grouping variable and the
second column contains the calculated summary statistics.

However, wrapping commands in this nested manner can make be confusing when 
trying to follow the chronology of commands. For this reason, the dplyr package provides
the use of the pipe operator (`%>%`), which improves the readability of your code.
The pipe operator takes the contents to the left of the it and passes those contents
into to the function on the right of it as if it was the first argument in that
function, making it easier to understand the chronology of commands when manipulating data. 

Returning back to our example, using the pipe operator as below returns the same results 
as the above line of code:

```{r, eval = FALSE}
group_by(growth_data, months) %>% summarize(mean(weight_kg))
```

It's even possible to use more than one pipe operator in succession. In this
case, we can instead use the pipe operator to pass the data into the `group_by`
function rather than provide it as the first argument of the function:

```{r, eval = FALSE}
growth_data %>% group_by(months) %>% summarize(mean(weight_kg))
```

We can also use line breaks after each pipe operator, which will automatically
be indented after the first line to represent the code as a single expression:

```{r, eval = FALSE}
growth_data %>%
  group_by(months) %>%
  summarize(mean(weight_kg))
```

To be clear, all four of the previous chunks of code give the same result. For 
space-saving purposes, we've suppressed the printed output. 

Here's one last thing, which we didn't cover in the guided practice. You can
provide a custom name to your newly created summary variable as follows:

```{r}
growth_data %>%
  group_by(months) %>%
  summarize(mean_weight_kg = mean(weight_kg))
```

Here I called my new variable `mean_weight_kg` for obvious reasons, but you are
free to provide whatever name you want. 

Try saving these results as an object in your global environment using the
assignment operator:

```{r}
growth_data_weights <-
  growth_data %>%
  group_by(months) %>%
  summarize(mean_weight_kg = mean(weight_kg))
```

Just as before, remember to save the work you've done. Try saving the script
you've been working in as `groupby_summarize.R` for later reference.
