---
title: "Module 2, Lesson 1: Simple linear regression"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

Open a new R script. At the top of this script, load the `tidyverse` and the
`healthds` package. 

```{r}
library(healthds)
library(tidyverse)
```

As you read about just before the Guided Practice, the `healthds` package 
contains several synthetic datasets for purposes of practicing statistical 
methods. Please refer that reading for additional details.

Loading the `healthds` package gives us access to these synthetic data using the 
following command:

(in the Guided Practice video, Dr. Boonstra mistakenly said that you first need to
use the `data("BPUrban300")` command -- that is not true.)

These data are stored as so-called data.frames and not tibbles, which is how
most of the data we've encountered so far are stored. For tibbles, R will
automatically trim the rows and columns that are printed, but for data.frame it
will not. The practical implication is that if you try to print these data to
the screen, it will be too much data for the screen. Instead, you can type
`View(BPUrban300)` in your console to open up a spreadsheet-like interface in
RStudio, or you can use the `head` function to view just the first few rows of
the data:

```{r}
head(BPUrban300)
```

Let's first inspect some of the variables in these data. A histogram of systolic
blood pressure (`systolic`) in these data can be constructed as such:

```{r}
ggplot(BPUrban300) + 
  geom_histogram(aes(systolic), bins = 30)
```

We can also inspect the relationship between systolic blood pressure and the
age of the person:

```{r}
ggplot(BPUrban300) + 
  geom_point(aes(age, systolic))
```

## Linear models in R

We will now fit a linear regression model, regressing `systolic` against `age`
to estimate the extent to which typical values systolic blood pressure change
with a person's age in these data. We will use the `lm` function in R to do so. 

The regression equation is specified via the `formula` argument in `lm`. A formula
starts with the name of a variable that is intended to be the outcome. Next comes
 a tilde (`~`) symbol. Finally, it ends with the name of a variable that is intended to be the covariate / predictor. So, our formula will be `systolic ~ age`. We also need to point `lm` to the dataset in which to "look" for these variables -- in our case that is `BPUrban300`. Putting this
 together, we can fit our regression model with the following:
 
```{r}
lm(formula = systolic ~ age, data = BPUrban300)
```

The output gives us just two things: the Call, which just gives back the code
that we gave it, and the Coefficients, which gives the intercept parameter
($\beta_0$ from Dr. Mukherjee's slides) and the slope parameter corresponding to
age ($\beta_1$). Importantly, this tells us that the expected systolic blood
pressure increases by about 0.5 points between any two people in our data who
are 1 year apart in age.

Congratulations on fitting your first linear regression. There is much more to
be done and learned with regard to regression. In future lessons, we will learn
more about how to conduct estimation and prediction.
 
 
 
 



