---
title: "Module 3, Lesson 2: From simple to multiple regression"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(out.extra='style="background-color: #999999; padding:7px; display: inline-block;"')
```


```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

```{r, message = F, warning = F}
library(healthds)
library(tidyverse)
```

Here is the code to fit the simple linear regression that we considered at the
end of Module 2, regressing systolic blood pressure against a dichotomized age:

```{r, eval = F}
BPUrban300 %>% 
  mutate(age_gt66 = case_when(age > 66 ~ "yes", 
                              age <= 66 ~ "no")) %>%
  lm(formula = systolic ~ age_gt66) %>%
  summary()
```

Now we will expand this model to additionally adjust for the body mass index (BMI, 
named `bmi` in the data) of the person (a number) as well as their smoking status (`smoking`, a TRUE/FALSE variable). To fit a multiple linear regression, we add more variable names to the right-hand
side of the `formula` argument. These variables names must all exist in the corresponding
dataset. 

```{r}
BPUrban300 %>% 
  mutate(age_gt66 = case_when(age > 66 ~ "yes", 
                              age <= 66 ~ "no")) %>%
  lm(formula = systolic ~ age_gt66 + bmi + smoking) %>%
  summary()
```

The most notable difference relative to a simple linear regression is that the
coefficients table now has four rows instead of two. The additional two rows
correspond to the additional regression coefficients that we've estimated. What
do these results tell us?

First, we see that there is a fairly small estimated association between BMI and
systolic blood pressure. BMI is measured in units of kg/m$^2$, and so this table
tells us that, for each kg/m$^2$ increase in BMI, we estimate that systolic
blood pressure increases by about 0.07 points, holding fixed the other variables
in the model (more on that in a moment). Keeping in mind that for most adults
BMI ranges from 15 to 40, for a total range of about 25. Thus, comparing a very
low BMI of 15 to a very large BMI of 40, we would only estimate the change in
the systolic blood pressure to be about 1.75 (or 25 times 0.07) points.
Moreover, this there is a great deal of uncertainty about this estimated
association: the standard error is about 0.33, resulting in a p-value of 0.829.
We would conclude from this analysis that this association is not statistically
significant, meaning that it is not inconsistent with the scenario where the
true association between BMI and systolic blood pressure is 0.

Let's turn to the phrase "holding fixed the other variables". What does this mean?
It is an important part of the interpretation of any multiple linear regression. 
Because we are adjusting for several variables at once, when we interpret any 
single association, we must assume that the other variables are not changing. 
So, for example, when we state that the expected systolic blood pressure is estimated
to increase by 0.07 points for someone whose BMI is 20 versus someone whose BMI is 19, 
we must also assume that these persons are in the same age group (holding fixed `age_gt66`)
and have the same smoking status (holding fixed `smokingTRUE`). We don't have to 
assume what age or smoking status they are, but just that they are the same. 

Now let's interpret the smoking association. Remember that smoking is a TRUE/FALSE
variable. Just as it did with `age_gt66`, R has chosen a reference category (FALSE, 
in this case) and assigned FALSE to take on values of 0 and TRUE to take on values
of 1. Thus, we interpret the estimated association of 2.83 to be the expected
increase in systolic blood pressure for someone who is a smoker compared to someone
who is not a smoker, holding fixed age and BMI. As before, there is quite a bit
of uncertainty surrounding this estimate, and the p-value tells us that this
estimated association is not inconsistent with something that we would observe
if the true association between smoking and systolic blood pressure was zero. 

One final comment about this model. The interpretation of the estimated
intercept has also changed, because we now must set to zero all of the
variables: the intercept is interpreted as the estimated systolic blood pressure
for someone whose age is in the reference category (younger than 66 years),
whose BMI is equal to zero, and whose smoking status is in the reference
category (non-smoker). Because it is biologically impossible to have a BMI of
zero, this intercept has no practical meaning.

### Fitting this same model to `BPUrban1000`

Recall that we have a separate dataset called `BPUrban1000`, comprised of 1000 observations from this same hypothetical target population. Let's look at what
our fitted model looks like when we regress upon this dataset with more than
triple the sample size:

```{r}
BPUrban1000 %>% 
  mutate(age_gt66 = case_when(age > 66 ~ "yes", 
                              age <= 66 ~ "no")) %>%
  lm(formula = systolic ~ age_gt66 + bmi + smoking) %>%
  summary()
```

Compared to the model fit to `BPUrban300`, the point estimates have changed
fairly substantially: the expected difference in systolic blood pressure between
the two age groups changed from 16.8 in the earlier model to 11.8 in this current
model. The estimated associations changed similarly for age and smoking status.
This change in point estimates is a consequence of sampling variability: the
estimated associations are random and will change from dataset to dataset.

Separate from this, we see that a consequence of increasing the sample size is
obtaining smaller standard errors: the standard errors are approximately 1/2 the
size of their counterparts from the `BPUrban300` model. Because we have this
gain in precision for estimating associations, the p-values are also closer to
zero. This highlights a subtle point regarding the concept of statistical
significance, namely that it is driven, in part, by sample size. The larger your
dataset, the more likely it is that you will identify a statistically
significant association (all other things being equal). This is not necessarily
a good thing, because statistical significance is not always the same as
practical significance, the latter referring to what the scientific consequences
of your findings are. For example, if you find that a particular dichotomous
risk factor is statistically significantly associated with a 0.09 point
change in average systolic blood pressure, that may not be a practically
significant finding. Even if you were confident that this was a causal relationship and
that policies could modify this risk factor in the population, the cost would
not justify such a small change. This relates to our last topic for this
independent guide, that of $R^2$, or the proportion of variance explained.

### Proportion of variance explained

At the very bottom of the output from our fitted model above, we see that 
the `Adjusted R-squared` is 0.054. Thus, this model does not explain 
a great deal of variability in the outcome. This may be surprising at first glance, 
given how statistically significant these associations are. But statistical significance
is not necessarily related to proportion of variability explained -- in fact this
is closely related to the distinction between statistical significance and practical
significance. 

To further highlight this point, let's inspect the `Adjusted R-squared` value for
the model regressing against dichotomized age alone in these data:

```{r}
BPUrban1000 %>% 
  mutate(age_gt66 = case_when(age > 66 ~ "yes", 
                              age <= 66 ~ "no")) %>%
  lm(formula = systolic ~ age_gt66) %>%
  summary()
```

That value is 0.040 (rounded to the 3rd digit), meaning that 4% of the variability in systolic blood pressure
is explained by dichotomized age alone. Adding both BMI and smoking status into
the model, we only explain an additional 1.4% of variation in that outcome, despite
them being both very statistically significantly related to systolic blood pressure. 
