---
title: "Module 2, Lesson 4: (even more) `dplyr`"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the fourth lesson of Module 2, you were introduced to the concept of joining
data. To practice using the joining functions, open the R project called
`Lessons.Rproj` and a blank R script. As before, you'll also need to
load the necessary packages into the current R session.

```{r}
library(healthds)
library(tidyverse)
```

## `full_join()`

Let's consider the `growth_data_aux` dataset contained in `healthds`. This
dataset contains measurements on the same infants as the `growth_data`.
However, there are four timepoints per infant in the `growth_data` while there
are only three timepoints per infant in the `growth_data_aux` dataset. The
`growth_data_aux` dataset also includes measurements of head circumference taken
at those timepoints.

```{r}
growth_data_aux
```

Notice that both datasets include the ID and month variables. These variables
serve as identifiers that help us identify the observations as well as match
and link data from different datasets via joining.

The `full_join` functions requires that you provide the names of the two datasets 
to join. By performing a full join of the datasets, we obtain a dataset of 
eight observations across five variables.

```{r}
full_join(growth_data, growth_data_aux)
```

Since head circumfrance measurements weren't taken at three months, R represents
these missing values as NA for "not available." In this case, the missing values
occur because a full join returns a row for every unique combination of the
identifier variables (e.g., ID and month variables).

## `left_join()` and `right_join()`

The `left_join` function ensures that every unique combination of the identifier
variables from the first dataset are returned in the joined data:

```{r}
left_join(growth_data, growth_data_aux)
```

This left join appears to give the same result as the full join we practiced earlier
since every unique combination of the identifier variables is captured in the first dataset.

On the other hand, the `right_join` function returns only the unique
combinations of the identifier variables that exist in the second dataset:

```{r}
right_join(growth_data, growth_data_aux)
```

Since the `growth_data_aux` dataset only contains measurements taken at three
timepoints, the joined data does not include the rows from the `growth_data` for
measurements taken at month three.

To check your understanding on the difference between these types of joins, try
to convince yourself why, for two datasets `x` and `y`, `left_join(x, y)` will 
yield the same result as `right_join(y, x)` (except that the order of the columns 
will be different). 

## `inner_join()`

The `inner_join` function returns the unique combinations of the identifier
variables that exist in both datasets:

```{r}
inner_join(growth_data, growth_data_aux)
```

For this example, this inner join gives an identical result as the right join we
did previously. However, it won't always be the
case that full and left joins return the same results, and the same is true for
right and inner joins. This behavior depends on the specific qualities of both
datasets.

Remember to save the work you've done. Try saving the script you've been working
in as `joining.R` for later reference.

## Help Pages

As discussed in the lecture, every R function ought to have a help page, which
is an incredibly useful tool for familiarizing yourself with a function.

To view the help page of a function, you can type a question mark and the name
of the function (e.g., `?pivot_wider`) in the console. The help page will appear
in the Help tab, which is typically in the lower right pane in RStudio.

Try pulling up the help page for the `pivot_wider` function. You'll notice that
this help page is comprised of a few different sections. The *Description* section
describes in plain language the intended behavior of the function. The *Usage*
section provides both the name of the function and all the possible function
arguments that you can specify. The *Arguments* section describes each argument in
more detail and the values that an argument accepts.

The *Examples* section provides example code of how the function is used. You can
try replicating the first example from the `pivot_wider` help page:

```{r}
fish_encounters
```

```{r}
fish_encounters %>%
  pivot_wider(names_from = station, values_from = seen)
```

(`fish_encounters` is a built-in dataset that is automatically available to you
when you install the `tidyverse`)

Now try pulling up the help page for the `select` function (i.e., `?select`).
You'll notice that this help page also gives an overview of the selection helper
functions (e.g., the `everything` function), which we discussed in the previous
lesson. Each of these helper functions also has their own help page. In this
case, the *Arguments* section isn't especially helpful since the arguments to
the `select` function are specific to a particular dataset in the form of
variable names.

Once again, you can try replicating the first example from the *Examples* section
of the help page:

```{r}
starwars %>% select(height)
```
