---
title: "Independent Guide: 2.4 - Joining Data"
output: 
  html_document:
  html_notebook:
editor_options: 
  chunk_output_type: inline
---

```{r klippy, echo = FALSE, include = TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy")
```

In the fourth lesson of Week 2, you were introduced to the concept of joining
data. To practice using the joining functions, open the R project called
`Lessons.Rproj` and a blank R script. Before we can begin, you'll also need to
load the necessary packages into the current R session.

```{r}
library(healthds)
library(tidyverse)
```

## full_join()

Let's consider the `growth_data_aux` dataset from the healthds package. This
dataset contains measurements on the same infants from the `growth_data`.
However, there are four timepoints per infant in the `growth_data` while there
are only three timepoints per infant in the `growth_data_aux` dataset. The
`growth_data_aux` dataset also includes measurements of head circumfrance taken
at those timepoints.

```{r}
growth_data_aux
```

Notice that both datasets include the ID and month variables. These variables
serve as identifiers that help us identify the observations, as well as match
and link data from different datasets via joining.

The `full_join` functions expects two datasets to join. By performing a full
join of the datasets, we obtain a data frame of eight observations across five
variables.

```{r}
full_join(growth_data, growth_data_aux)
```

Since head circumfrance measurements weren't taken at three months, R represents
these missing values as NA for "not available." In this case, the missing values
occur because a full join returns a row for every unique combination of the
identifier variables (e.g., ID and month variables).

## left_join() and right_join()

The `left_join` function ensures that every unique combination of the identifier
variables from the first dataset are returned in the joined data.

```{r}
left_join(growth_data, growth_data_aux)
```

This left join appears the same as the full join since every unique combination
of the identifier variables is captured in the first dataset.

On the other hand, the `right_join` function returns only the unique
combinations of the identifier variables that exist in the second dataset.

```{r}
right_join(growth_data, growth_data_aux)
```

Since the `growth_data_aux` dataset only contains measurements taken at three
timepoints, the joined data does not include the rows from the `growth_data` for
measurements taken at month three.

## inner_join()

The `inner_join` function returns the unique combinations of the identifier
variables that exist in both datasets.

```{r}
inner_join(growth_data, growth_data_aux)
```

This inner join appears the same as the right join since the measurements taken
at month three do not exist in both datasets. However, it won't always be the
case that full and left joins return the same results, and the same is true for
right and inner joins. This behavior depends on the specific qualities of both
datasets.

Remember to save the work you've done. Try saving the script you've been working
in as `joining.R` for later reference.

## Help Pages

As discussed in the lecture, every R function ought to have a help page, which
is an incredibly useful tool for familiarizing yourself with a function.

To view the help page of a function, you can type a question mark and the name
of the function (e.g., `?pivot_wider`) in the console. The help page will appear
in the Help tab, which is typically in the lower right pane in RStudio.

Try pulling up the help page for the `pivot_wider` function. You'll notice that
this help page is comprised of a few different sections. The Description section
describes in plain language the intended behavior of the function. The Usage
section provides both the name of the function and all the possible function
arguments that you can specify. The Arguments section describes each argument in
more detail and the values that an argument accepts.

The Examples section provides example code of how the function is used. You can
try replicating the first example from the `pivot_wider` help page.

```{r}
fish_encounters
```

```{r}
fish_encounters %>%
  pivot_wider(names_from = station, values_from = seen)
```

Now try pulling up the help page for the `select` function (i.e., `?select`).
You'll notice that this help page also gives an overview of the selection helper
functions (e.g., the `everything` function), which we discussed in the previous
lesson. Each of these helper functions also has their own help page. In this
case, the Arugments section isn't especially helpful since the arguments to
the `select` function are specific to a particular dataset in the form of
variable names.

Once again, you can try replicating the first example from the Examples section
of the help page.

```{r}
starwars %>% select(height)
```
